Fast and Flexible Binary Instrumentation for the Kernel

### 2013-12-05 VEE-14/paper.tex
Decided to elaborate on return address transparency. Specifically, on why Granary does not end basic blocks at CALL instructions.

Associated with each basic block is meta-data that describes the basic block. Basic block meta-data includes: \begin{itemize}
	\item The length in bytes of the original and translated basic blocks, and the address of the first instruction in the original basic block and the instrumented basic block. This information helps provide a rich debugging interface for instrumentation tool authors. Granary is unique among DBT tools in that the debugging experience is actually pleasant. For example, Granary has a built-in debugging mechanism for stepping back in time through previously executed basic blocks, as well as querying past register state.
	\item Linux kernel-specific exception table meta-data about instructions contained within the current basic block. This kernel-specific meta-data is used by Granary to implement transparent page fault exception handling (\Cref{sec:interrupt_return_address_transparency}).
	\item Ranges of instructions that must be executed atomically. Granary uses this information to implement selective interrupt delaying (\Cref{sec:interrupt_delay}).
	\item The allocator used to allocate the instructions of the basic block. Granary uses this information to implement its trace-allocation optimization (\Cref{sec:trace_alloc}).
	\item The instrumentation policy used to instrument the basic block. Granary uses this information to allow instrumentation tools to define policy-specific interrupt handling routines. For example, \toolname{RCUdbg} implements a read-side policy-specific interrupt handling routine that reports an error if a code executing within a read-side critical section is interrupted.
\end{itemize} %
%Granary uses basic block meta-data for trace allocation (\Cref{sec:trace_alloc}), interrupt delaying (\Cref{sec:interrupt_delay}), and policy-specific interrupt handling.

Basic block meta-data can be also be extended by instrumentation tools. This makes developing instrumentation tools that record per-basic block statistics very easy to make. For example, Granary's control-flow graph tool, \toolname{cfg}, extends the basic block meta-data to track per-block and per-edge execution counts. This data is useful because it provides precise code coverage information (block and branch coverage), as well as useful data for profile-guided optimization. This meta-data is specified by \toolname{cfg} and automatically allocated for \toolname{cfg} by Granary.

@@

\subsubsection{Function return addresses}\label{sec:return_address_transparency}
Unlike most DBT frameworks, Granary includes \texttt{call} instructions into basic blocks, which exposes code cache addresses to instrumented module code in the form of return addresses. The benefit of including \texttt{call} instructions is better return-address prediction by the processor and longer basic blocks, which provide more code visibility to instrumentation tools at instrumentation time.

@@

After Granary bootstraps on the module initialization process, \emph{attaching} occurs in one of three ways: \begin{enumerate}
	\item {\bf Implicit attaching:} the kernel returns to instrumented module code in the code cache. This occurs when instrumented moduel code is interrupted, or when the module invokes a kernel function (detaching) and that function returns to the instrumented module code (attaching).
	\item {\bf Fast attaching:} the kernel invokes a wrapped module function. The first instance of fast attaching for any kernel module is when the kernel's module loader invokes the instrumented module initializer function.
	\item {\bf Slow attaching:} the kernel invokes unwrapped module code. If the kernel executes a module function which was passed to the kernel in a type-unsafe manner, then the processor will raise a fault because Granary uses hardware page protection to prevent module code from being executed. Granary handles these faults by returning execution to the instrumented version of the faulting module code.
\end{enumerate}

@@

>>>>

Associated with each basic block is meta-data that describes the basic block. Basic block meta-data includes: \begin{itemize}
	\item The length in bytes of the original and translated basic blocks, and the address of the first instruction in the original basic block and the instrumented basic block. This information helps provide a rich debugging interface for instrumentation tool authors. Granary is unique among DBT tools in that the debugging experience is actually pleasant. For example, Granary has a built-in debugging mechanism for stepping back in time through previously executed basic blocks, as well as querying past register state.
	\item Linux kernel-specific exception table meta-data about instructions contained within the current basic block. This kernel-specific meta-data is used by Granary to implement transparent page fault exception handling (\Cref{sec:interrupt_return_address_transparency}).
	\item Ranges of instructions that must be executed atomically. Granary uses this information to implement selective interrupt delaying (\Cref{sec:interrupt_delay}).
	\item The allocator used to allocate the instructions of the basic block. Granary uses this information to implement its trace-allocation optimization (\Cref{sec:trace_alloc}).
	\item The instrumentation policy used to instrument the basic block. Granary uses this information to allow instrumentation tools to define policy-specific interrupt handling routines. For example, \toolname{RCUdbg} implements a read-side policy-specific interrupt handling routine that reports an error if a code executing within a read-side critical section is interrupted (\Cref{sec:rcudbg}).
\end{itemize} %
%Granary uses basic block meta-data for trace allocation (\Cref{sec:trace_alloc}), interrupt delaying (\Cref{sec:interrupt_delay}), and policy-specific interrupt handling.

Basic block meta-data can be also be extended by instrumentation tools. This makes developing instrumentation tools that record per-basic block statistics very easy to make. For example, Granary's control-flow graph tool, \toolname{cfg}, extends the basic block meta-data to track per-block and per-edge execution counts (\Cref{sec:cfg}). This data is useful because it provides precise code coverage information (block and branch coverage), as well as useful data for profile-guided optimization. This meta-data is specified by \toolname{cfg} and automatically allocated for \toolname{cfg} by Granary.

@@

\subsubsection{Function return addresses}\label{sec:return_address_transparency}
Unlike most DBT frameworks, Granary's basic blocks do not terminate at \texttt{call} instructions. This exposes code cache addresses to instrumented module code in the form of return addresses stored on the runtime call stack. The benefit of including \texttt{call} instructions are: \begin{description}[labelindent=\parindent]

	\item[i) Zero overhead procedure returns.] Unlike Granary, DRK emulates procedure calls and returns. DRK emulates \texttt{call} instructions by pushing the intended native return address onto the runtime call stack, followed by a direct direct jump to the next (instrumented) basic block \cite{DRK}. This approach enables return address transparency, but introduces additional overheads on procedure returns, which must do costly indirect branch lookups to find the intended return target.

Btkernel's approach is similar in spirit to Granary's; however, btkernel ends basic blocks at \texttt{call} instructions. This introduces unnecessary control-flow on the return path from procedures. That is, btkernel's ``call/ret optimization" ostensibly enables zero-overhead procedure returns by leaving \texttt{ret} instructions as-is; however, the benefits are more apparent than real. To transfer control to the next logical basic block after a procedure return, btkernel's \texttt{ret}s first transfer control to the end of the basic block that performed the \texttt{call}, and then \texttt{jmp} to the next basic block. Granary avoids this unnecessary \texttt{jmp} because basic blocks do not end at \texttt{call} instructions.

In some cases, emulating DRK's return-address lookup mechanism is desirable. For example, if Granary attaches while some program's execution is underway (and so the runtime call stack contains native return addresses), then a direct \texttt{ret} back to one of the native return addresses might be undesirable (e.g. if the tool wants comprehensive instrumentation of all code). In such cases, a tool can partially specialize instrumentation according to whether or not the return address might be in the code cache. Partial specialization is enabled by \emph{policy properties}, which are transparently inherited along with normal instrumentation policy information. In this case, the policy property can be used to tell Granary to emulate \texttt{ret} instructions when the return address is potentially a native address, and directly \texttt{ret} when the return address is known to be in the code cache.\footnote{This feature is used when Granary is configured to instrument user space programs. User space Granary depends on the Linux dynamic loader to inject Granary into a running process. When injected, a Granary takeover function is executed on an already active native runtime call stack. Granary's takeover procedure transparently emulates returns, thus taking over the previously fully-native call stack. This feature allows instrumentation tools to focus on their particular objectives and not the intricacies of thread takeover.}

%performing a \texttt{ret} requires an additional \texttt{jmp} to 
%Neither of DRK's nor btkernel's approaches enable native-like return address prediction by the CPU.
%and then jumping to the next basic block. This in turn . Btkernel's approach to procedure calls introduces similar inefficiencies, despite the fact that does not provide return address transparency, because it also emulates procedure calls, but by manually pushing the instrumented return address onto the stack instead of the native one.
%Granary's approach improves on DRK and btkernel by allowing 
%Better return-address prediction by the processor.

	\item[iii) Improved visibility.] Not ending basic blocks on \texttt{call} instructions results in longer basic blocks, which in turn provide additional visibility to tools at instrumentation time into what will be executed by the current thread. \toolname{RCUdbg} benefits from this additional visibility because it can observe the end of read-side critical sections (\texttt{call}s to \texttt{rcu\_read\_unlock}), and optimistically assume that functions executing before the read-side critical section ends will execute within the read-side critical section.

	\item[ii) Zero-overhead attaching.] (\Cref{sec:implicit_attach})

	%longer basic blocks, which provide more code visibility to instrumentation tools at instrumentation time.
\end{description}

@@

\label{sec:implicit_attach}
After Granary bootstraps on the module initialization process, \emph{attaching} occurs in one of three ways: \begin{enumerate}
	\item {\bf Implicit attaching:} the kernel returns to instrumented module code in the code cache. This occurs when instrumented moduel code is interrupted, or when the module invokes a kernel function (detaching) and that function returns to the instrumented module code (attaching).
	\item {\bf Fast attaching:} the kernel invokes a wrapped module function. The first instance of fast attaching for any kernel module is when the kernel's module loader invokes the instrumented module initializer function. 
	\item {\bf Slow attaching:} the kernel invokes unwrapped module code. If the kernel executes a module function which was passed to the kernel in a type-unsafe manner, then the processor will raise a fault because Granary uses hardware page protection to prevent module code from being executed. Granary handles these faults by returning execution to the instrumented version of the faulting module code.
\end{enumerate}

@@

\section{Application: Control-Flow Graphs}\label{sec:cfg}
